---
title: "Topographic Change caused by People"
author: "Matthew Ross"
date: "9/23/2019"
output: html_document
editor_options: 
  chunk_output_type: console
---


# Analyzing topographic change in the USA

```{r setup, include=FALSE}
library(tidyverse)
library(sf)  #Spatial datasets
library(mapview) #interactive mapping
library(raster) #Raster analysis
library(elevatr) #National Elevation Dataset downloader. 
library(ggthemes) #for plotting
library(ggplot2) #for plotting

knitr::opts_chunk$set(echo = TRUE)
```


## USGS analysis of topographic change

Analysis background can be found here: https://www.usgs.gov/land-resources/eros/topochange


### Download the data


```{r, eval=F}
#Create a new folder 
dir.create('data')

#Download files
download.file(url='https://edcintl.cr.usgs.gov/downloads/sciweb1/shared/topo/downloads/Topochange/topo_change_polygons_Shapefile.zip',destfile='data/topo_change.zip')

# List files in the data directory
#list.files('data')

#Unzip the dataset
unzip('data/topo_change.zip',exdir='data/topo_change')

#Check that the unzipping worked
list.files('data/topo_change')

```

### Check out the data

```{r}

topo_change <- st_read('data/topo_change/topo_change_polygons.shp')

#Checkout the column names
#names(topo_change)

biggest_area <- topo_change %>%
  arrange(desc(AREA_SQ_KM)) %>%
  slice(1:10)

# Map all mines
mapview(biggest_area)
```

### Checkout mine in Arizona

```{r}
# Subset to just Arizona
az_mine <- biggest_area %>%
  filter(QUADNAME == 'esperanza_mill_AZ')

#Check that it is the right site
mapview(az_mine)
```



## Getting our own elevation datasets

### Download pre-mining DEM 

```{r}

#Check projection of az_mine
#st_crs(az_mine)

az_raster_before <- get_elev_raster(az_mine,z=12)

#Look at the structure of the data
#str(az_raster_before)

#Summary of the data
#summary(az_raster_before)

plot(az_raster_before)

```


### Download post-mining DEM

Data comes from [SRTM](https://www2.jpl.nasa.gov/srtm/) 

```{r,eval=F}
#Extract centroid of mine polygon
az_center <- st_centroid(az_mine) %>%
  st_transform(4326)
#Map it to check
mapview(az_center)

#Get coords 
coords <- st_coordinates(az_center)


getData('SRTM',lon=coords[1],lat=coords[2],
        path='data')

```


### Read in SRTM data
```{r}
az_raster_after <- raster('data/srtm_14_06.tif')

```

## Looking at Elevation change before and after mining

### Clip SRTM data to NED data

```{r}
#Check resolution of rasters
# res(az_raster_after)
# res(az_raster_before)

#Check the projection
# projection(az_raster_after)
# projection(az_raster_before)


#Reproject small raster to larger raster projection
az_raster_before_reproj <- projectRaster(az_raster_before,
                                        az_raster_after)

#Trim whitespace
az_raster_before_trim <- trim(az_raster_before_reproj) 

#Crop to match extents
az_raster_after_clip <- crop(az_raster_after,az_raster_before_trim)

#Plot clipped dataset
plot(az_raster_after_clip)
```


### Taking the difference and plotting it

```{r}
# Take the elevation difference
dem_diff <- az_raster_before_trim - az_raster_after_clip

plot(dem_diff)

```


# In-class work

## 1) What is the maximum/minimum elevation change caused by mining?

You may want to check out the command maxValue and minValue

```{r}
#YOUR CODE HERE

#Minimum value caused by mining
minValue(dem_diff)

#Maximum value caused by mining
maxValue(dem_diff)

#Positive elevation change is within the map area, but not at the mining site. Positive elevation change from the maxValue() function could still be reflecting mining activity, such as piles of unused material. 

```


## 2) What is the background elevation "error" between NED and SRTM? 

The national elevation dataset (NED) and the Shuttle 
Radar Topography Mission are extremely different ways to
estimate elevatin. NED is derived from thousands of 
compiled elevation datasets that can be collected on the
ground, from LiDAR or other ways, while SRTM is collected
with radar from space. As a result, we naturally expect
there to be some background differences in these 
two DEMs, so what is the average difference in these DEMs
in parts of the landscape that have not undergone substantial 
elevation change? 

### 2a) Buffer the az_mine object by 5 kilometers using the sf::st_buffer command

```{r}

#Your code here
#Creating a 5 km buffer. Remember: az_mine units is in meters
az_buff <- st_buffer(az_mine, dist = 5000)

#mapview to check if buffer worked
mapview(az_mine) + mapview(az_buff)

```


### 2b) Use the the st_transform command to transform the 
projection of your buffered az_mine dataset to the 
same projection as the dem_diff raster dataset

hint: you may want to explore st_crs to help you here

```{r}
#project az_buff to dem_diff projection
az_buff_p <- st_transform(az_buff, crs = st_crs(dem_diff))

#check the projection to make sure it works
#st_crs(dem_diff)
#st_crs(az_buff_p)

```


### 2c) Use the raster::mask command to mask out all the 
areas inside of the buffered and reprojected az_mine dataset.

hint: look at the mask command and try to understand what
the inverse option does

```{r}
#Your code here

#mask NED topography
az_NED_mask <- mask(az_raster_before_trim, az_buff_p, inverse = TRUE)
#mapview(az_NED_mask) # just to check

#mask STRM topography
az_STRM_mask <- mask(az_raster_after_clip, az_buff_p, inverse = TRUE)
#mapview(az_STRM_mask) # just to check
```


### 2d) Use the summary command to look at the summary of
differences between these two DEMs in areas where there has
been no mining

```{r}
#find the difference between the two topographies
az_mask_diff <- az_NED_mask - az_STRM_mask

#summarize differences
summary(az_mask_diff)

#view differences just to see
#mapview(az_mask_diff)
#plot(az_mask_diff)
```


## 3) What does the shape of the "error" distribution look like
in areas that have not been mined versus areas that have
been mined? 

To look at distributional changes, we may want to convert this
raster dataset back into a data.frame or tibble type shape so 
that we can take advantage of ggplots plotting capabilities and 
make a plot of the two dems.

### 3a) Mask the raster dem_diff dataset to the exact outlines of the
USGS az_mine dataset

```{r}
#mask the dem by the mine outline + transform mine to dem_diff crs
dem_diff_mine <- mask(dem_diff, 
                      st_transform(az_mine, st_crs(dem_diff)))

#plot(dem_diff_mine)

```


### 3b) Use the getValues command to extract the elevation
values from both the unmined and mined regions of your rasters. 
Label these vectors as `unmined_dem` and `mined_dem`. 

```{r}
#get the unmined area of dem_diff
dem_diff_unmine <- mask(dem_diff, 
                        st_transform(az_mine, st_crs(dem_diff)), 
                        inverse = TRUE)
#plot(dem_diff_unmine)

#create two vectors of mined and unmined dem values
mined_dem <- getValues(dem_diff_mine)
unmined_dem <- getValues(dem_diff_unmine)

#summary(mined_dem) # lots of NAs
#summary(unmined_dem) # not as many NAs
```


### 3c) Use ggplot and the command geom_density() to look at
the differences in mined and unmined regions of Arizona.

```{r}
# Here I'm just making the table for you

unmined_type <- rep('unmined',length(unmined_dem))
mined_type <- rep('mined',length(mined_dem))
mined_table <- tibble(dem_values = c(unmined_dem,mined_dem),
                      type = c(unmined_type,mined_type)) 
                      

# plot the density plots
ggplot(data = mined_table, aes(x = dem_values, colour = type)) + 
  geom_density(size = 1) + 
  ylab('Density') + 
  xlab('Difference DEM Values (m)') + 
  theme_few() + 
  scale_color_few()+
  theme(legend.position=c(0.1,0.85))

#DEM differences in the unmined area are centered around zero. Differences in the mined area are centered around approx. -40 m
                    
```

